A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-20 tex/body.tex
more

\newcommand\DISC[1]{t \teq \const{C}_{#1}\bigl(\const s^1_{#1}( t ), \ldots,\const s^{n_{#1}}_{#1}( t )\bigr)}
%\newcommand\DISC[1]{\const{d}_{#1}(t)}

@@

\hyphenation{data-type data-types co-data-type co-data-types isa-belle sledge-hammer
non-redun-dant}

@@

\begin{rep}\ourparagraph{Related Work.} Despite\end{rep}%
\begin{conf}Despite\end{conf} its age,
Barrett et al.\ \cite{barrett-et-al-2007} still provides a good account of
related work. Since then, datatypes have been added to the SMT solver
Z3 \cite{de-moura-bjoerner-2008} \begin{rep}in unpublished work by Leonardo de Moura \end{rep}and to a SPASS-like %prototypical
superposition prover \begin{rep}by Daniel Wand\end{rep} \cite{wand-2014}.
Closely related are the automatic structural
induction in both kinds of provers \cite{kersani-peltier-2013,cruanes-201x,wand-weidenbach-201x,reynolds-kuncak-2015},
the (co)datatype and (co)induction support in Dafny \cite{leino-moskal-2014},
and the (semi-)decision procedures for datatypes
implemented in Leon \cite{suter-et-al-2011} and RADA \cite{pham-et-al-2013}.
%
Datatypes are supported by the higher-order model finder Refute
\cite{weber-2008}. Its successor, Nitpick \cite{blanchette-nipkow-2010}, can
also generate models involving cyclic codatatype values. The $\mu$-notation is
inspired by the $\mu$-calculus
%\begin{conf}\cite{kozen-1983}.\end{conf}%
%\begin{rep}\cite[etc.]{kozen-1983,endrullis-et-al-2011}.\end{rep}%
\cite[etc.]{kozen-1983,endrullis-et-al-2011}.

@@

%\ourparagraph{Integration into {\relax\XXXL($T\negvthinspace$)}.}
{\looseness=-1
The decision procedure is implemented as a \relax{theory solver} of CVC4, that is, a
specialized solver for determining the satisfiability of conjunctions of
constraints for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a set of
input clauses $F$ in conjunctive normal form, the \XXXL($T$) procedure
incrementally builds partial assignments of
truth values to the atoms of $F$ such that no clause in $F$ is falsified. We can regard such a
partial assignment as a set $M$ of true literals.
% where $a \in M$ for all atoms $a$ that
%it assigned the value $\top$ and $\neg\; a \in M$ for all atoms that it assigns
%the value $\bot$.
%at a high level,
By a variant of the Nelson--Oppen method \cite{jovanovic2011sharing,nelson-oppen-1979},
each $T_i$-solver %for $i = 1, \ldots, l$  n?
takes as input the union $M_i$
of (1)~the purified form of $T_i$-constraints occurring in $M$, where fresh
variables replace terms containing symbols not belonging to~$T_i$;
(2)~additional (dis)equalities between variables of types not belonging to
$T_i$. Each $T_i$-solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\; C$ is added to $F\!$, adds a clause to $F\!$, or does nothing.
When $M$ is a complete assignment for $F\!$, a theory solver can choose to do
nothing only if $M_i$ is indeed $T_i$-satisfiable.

@@

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the theory files from the Isabelle distribution (Distro, 879~goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We added
two unpublished theories by Peter Gammie and Andreas Lochbihler (\gandl, 317
goals), about Bird and Stern--Brocot trees. To exercise the datatype
support, the benchmarks were complemented by theories
about lists and trees. The theories were selected before running the
experiments. The experimental data are publicly available.%
\footnote{\url{http://www21.in.tum.de/~blanchet/codp_data.tgz}}

@@

We introduced a decision procedure for the ground theory of datatypes and
codatatypes. Our main
contribution has been the support for codatatypes. Both the metatheory and
the implementation in CVC4 rely on $\mu$-terms to represent cyclic values,
a complication specific to codatatypes.
% are not very
% difficult per se, but there are tricky corner cases to take care of if we care
% about completeness (which we do).
%% alliteration
The empirical results on Isabelle benchmarks
confirm that CVC4's new capabilities improve the state of the art.

This work is part of a program that aims at enriching automatic provers
with high-level features and at reducing the gap between automatic and
interactive theorem proving. As future work, it would be useful to implement
proof reconstruction for (co)data\-type inferences in Isabelle.
CVC4 cound be interfaced for generating counterexamples in proof assistants, as
an alternative to Nitpick; in this context, the acyclicity and uniqueness
rules are crucial to exclude spurious countermodels. Finally, it might be
worthwhile to extend SMT solvers with dedicated support for (co)recursive
functions.

>>>>

\newcommand\DISC[1]{t \teq \const{C}_{#1}\negvthinspace\bigl(\const s^1_{#1}( t ), \ldots,\const s^{n_{#1}}_{#1}( t )\bigr)}
%\newcommand\DISC[1]{\const{d}_{#1}(t)}

@@

\hyphenation{data-type data-types co-data-type co-data-types isa-belle sledge-hammer
non-redun-dant counter-examples counter-example}

@@

\begin{rep}\ourparagraph{Related Work.} Barrett\end{rep}%
\begin{conf}Barrett\end{conf}
et al.\ \cite{barrett-et-al-2007} provide a good account of
related work. Since then, datatypes have been added to the SMT solver
Z3 \cite{de-moura-bjoerner-2008} \begin{rep}in unpublished work by Leonardo de Moura \end{rep}and to a SPASS-like %prototypical
superposition prover \begin{rep}by Daniel Wand\end{rep} \cite{wand-2014}.
Closely related are the automatic structural
induction in both kinds of provers \cite{kersani-peltier-2013,cruanes-201x,wand-weidenbach-201x,reynolds-kuncak-2015},
the (co)datatype and (co)induction support in Dafny \cite{leino-moskal-2014},
and the (semi-)decision procedures for datatypes
implemented in Leon \cite{suter-et-al-2011} and RADA \cite{pham-et-al-2013}.
%
Datatypes are supported by the higher-order model finder Refute
\cite{weber-2008}. Its successor, Nitpick \cite{blanchette-nipkow-2010}, can
also generate models involving cyclic codatatype values. The $\mu$-notation is
inspired by the $\mu$-calculus
%\begin{conf}\cite{kozen-1983}.\end{conf}%
%\begin{rep}\cite[etc.]{kozen-1983,endrullis-et-al-2011}.\end{rep}%
\cite[etc.]{kozen-1983,endrullis-et-al-2011}.

@@

%\ourparagraph{Integration into {\relax\XXXL($T\negvthinspace$)}.}
The decision procedure is implemented as a \relax{theory solver} of CVC4, that is, a
specialized solver for determining the satisfiability of conjunctions of
constraints for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a set of
input clauses $F$ in conjunctive normal form, the \XXXL($T$) procedure
incrementally builds partial assignments of
truth values to the atoms of $F$ such that no clause in $F$ is falsified. We can regard such a
partial assignment as a set $M$ of true literals.
% where $a \in M$ for all atoms $a$ that
%it assigned the value $\top$ and $\neg\, a \in M$ for all atoms that it assigns
%the value $\bot$.
%at a high level,
By a variant of the Nelson--Oppen method \cite{jovanovic2011sharing,nelson-oppen-1979},
each $T_i$-solver %for $i = 1, \ldots, l$  n?
takes as input the union $M_i$
of (1)~the purified form of $T_i$-constraints occurring in $M$, where fresh
variables replace terms containing symbols not belonging to~$T_i$;
(2)~additional (dis)equalities between variables of types not belonging to
$T_i$. Each $T_i$-solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\, C$ is added to $F\!$, adds a clause to $F\!$, or does nothing.
When $M$ is a complete assignment for $F\!$, a theory solver can choose to do
nothing only if $M_i$ is indeed $T_i$-satisfiable.

@@

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the theory files from the Isabelle distribution (Distro, 879~goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We added
two unpublished theories by Peter Gammie and Andreas Lochbihler (\gandl, 317
goals), about Bird and Stern--Brocot trees. To exercise the datatype
support, the benchmarks were complemented by theories
about lists and trees. The theories were selected before running the
experiments. The experimental data are publicly available.%
\footnote{\url{http://lara.epfl.ch/~reynolds/CADE2015-cdt/}}
%http://www21.in.tum.de/~blanchet/codp_data.tgz}}

@@

We introduced a decision procedure for the ground theory of datatypes and
codatatypes. Our main
contribution has been the support for codatatypes. Both the metatheory and
the implementation in CVC4 rely on $\mu$-terms to represent cyclic values.
Although this aspect is primarily motivated by codatatypes, it enables
a uniform account of datatypes and codatatypes---in particular,
the acyclicity rule for datatypes exploits $\mu$-terms to detect cycles.
% are not very
% difficult per se, but there are tricky corner cases to take care of if we care
% about completeness (which we do).
%% alliteration
The empirical results on Isabelle benchmarks
confirm that CVC4's new capabilities improve the state of the art.

This work is part of a program that aims at enriching automatic provers
with high-level features and at reducing the gap between automatic and
interactive theorem proving. As future work, it would be useful to implement
proof reconstruction for (co)data\-type inferences in Isabelle.
CVC4 could also be interfaced for generating counterexamples in proof assistants, as
an alternative to Nitpick; in this context, the acyclicity and uniqueness
rules are crucial to exclude spurious countermodels. Finally, it might be
worthwhile to extend SMT solvers with dedicated support for (co)recursive
functions.
